---
title: "Step AIC"
author: "Alan T. Arnholt"
date: "9/16/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(MASS)
names(Boston)
cor(Boston)
stepAIC(lm(medv ~ 1, data = Boston), scope = (~crim+zn + indus + chas + nox + rm + age +dis + rad + tax + ptratio + black +lstat), direction = "forward", test = "F")
### OR
null <- lm(medv ~ 1, data = Boston)
full <- lm(medv ~ ., data = Boston)
stepAIC(null, scope = list(lower = null, upper = full), direction = "forward", test = "F")
```
```{r}
mod <- lm(medv ~ lstat + rm + ptratio + dis, data = Boston)
summary(mod)
# 
X <- model.matrix(mod)
XTXI <- solve(t(X)%*%X)
XTXI
# But best to get XTXI using cov.unscaled
XTXI <- summary(mod)$cov.unscaled
XTXI
betahat <- XTXI%*%t(X)%*%Boston$medv
betahat
MSE <- summary(mod)$sigma^2
vcmat <- MSE*XTXI
vcmat
se <- diag(vcmat)^.5
ts <- betahat/se
pvalues <- pt(abs(ts), 501, lower = FALSE)*2
pvalues
CI <- betahat[2] + c(-1, 1)*qt(.975, 501)*se[2]
CI
confint(mod)
```

```{r}
par(mfrow = c(2, 2))
plot(mod)
par(mfrow = c(1,1))
```

```{r}
mod2 <- lm(medv ~ . - age - indus, data = Boston)
summary(mod2)
plot(mod2)
```

```{r}
library(car)
residualPlots(mod)
modM <- lm(medv ~ lstat + rm + ptratio + dis + I(rm^2) + I(lstat^2), data = Boston)
residualPlots(modM)
plot(modM)
summary(modM)
```

```{r}
residualPlots(mod2)
mod3 <- update(mod2, .~. + I(lstat^2) + I(rm^2))
mod3
summary(mod3)
residualPlots(mod3)
```

## Conceptual and Computational

```{r}
airline <- readr::read_csv("../Data/airline.csv")
airline
```
* Fit a linear model by regressing `fare` onto `miles`.

```{r}
mod <- lm(fare ~ miles, data = airline)
summary(mod)
```

* Find $R^2$

The $R^2$ value is `r summary(mod)$r.squared`.

* What does $R^2$ mean in this context? 

`r round(summary(mod)$r.squared*100, 2)`% of the variability in airfares is explained by changes in distance.

* What is the slope of the regression line?

`r summary(mod)$coef[2, 1]`

* What is the $y$-intercept?

`r summary(mod)$coef[1, 1]`

* What is the equation of the linear model?

$\widehat{fare} = `r summary(mod)$coef[1, 1]` + `r summary(mod)$coef[2, 1]` \times \text{miles}$

* Estimate the fare for a 200 mile flight

```{r}
predict(mod, newdata = data.frame(miles = 200))
```

* Estimate the fare for a 2000 mile flight

```{r}
predict(mod, newdata = data.frame(miles = 2000))
```

* Explain what the y intercept means in this context.

The model says a flight of zero miles will cost $177.21. This could be fixed costs of air travel.

* Explain what the slope means in this context.

An increase of 1 mile means an increase of 7.86 cents.

* The fare to fly to Las Vegas, 1719 miles from Atlanta, is $252. What is the residual?

```{r}
yhat <- predict(mod, newdata = data.frame(miles = 1719))
yhat
(resid <- 252 - yhat)
```

* In general a positive residual means...

The model has underpredicted the actual value

* In general a negative residual means...

The model has overpredicted the actual value

* Is a linear model appropriate for estimating airfare from distance flown? Why or Why not?

```{r}
library(ggplot2)
ggplot(data = airline, aes(x = miles, y = fare)) + 
  geom_point() + 
  theme_bw() + 
  geom_smooth(se = FALSE, method = "lm")
plot(mod, which = 1)
```

Yes, the scatterplot shows no curve, and the residual plot has no outliers, fan shape, or curve.

* How strong is the model? Explain.

The model is moderately strong according to an $R^2$ value of `r round(summary(mod)$r.squared*100,2)`%.

* Identify any outliers. Why are they unusual?

No outliers.

* If you want to predict the distance of a flight from the fare, how does the model change?

```{r}
mod2 <- lm(miles ~ fare, data = airline)
summary(mod2)
```

$\widehat{\text{miles}} = -644.29 + 6.131 \times \text{fare}$

Use the fare calculated for a 2000 mile flight ($334.4526) to find out how far the model above suggest you could fly.

```{r}
predict(mod2, newdata = data.frame(fare = 334.4526))
```

